Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-02-09 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

The computational domain with permeability distribution of 10$^{-4}\le\mathbf{K}\le$10$^{5}$  milidarcy (Fig.~\ref{fig:3DChannel_PermMesh})  was designed to naturally capture preferential flow pathways through channels of large absolute permeability. Pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$ from the left-hand side face. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid, \ie $S_{nw}(t=0)=0.80$, which was continuously displaced by the wetting fluid as the simulation progress.

As demonstrated in the preliminary model validation (Section~\ref{section:results_initial_model_validation}), crossflow towards the upper region of the initial part of the domain is due to boundary pressures and diffusivities (both at different vertical layers), \ie caused by large permeability gradients (Fig.~\ref{fig:3DChannel_sat}). As the simulation progresses, non-wetting fluid is driven from regions with lower fluid resistance (central channel), although adjacent region (with low permeability) remains stagnant (Fig.~\ref{fig:3DChannel_planes}).

\medskip
Flow pathways are clearly indicated in Fig.~\ref{fig:3DChannel_planes}, where Darcy velocity vectors overlapped with wetting phase fluid saturation are shown. Fluid displacement occurs mainly in regions of relatively larger permeability, leaving other parts of the domain (\ie regions with low permeability distribution) with little or no momentum, indicating that at these regions the non-wetting fluid (\ie fluid 2) are not effectively displaced.

\medskip
In Sections \ref{section:results_homo_hete}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. Figure~\ref{fig:3DChannel_sat} shows preferential flow pathways at various instances of time. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces (Fig.~\ref{fig:3DChannel_Isosurf}).

\medskip
A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}, with resolution ranging from 235242 to 643064 of tetrahedral \PNDG[1]{1} element-pairs (Fig.~\ref{fig:3DChannel_severalfields_adapt}a). Distinct preferable flow pathways can be shown (Fig.~\ref{fig:3DChannel_severalfields_adapt}c), in particular through the central channel (see streamlines in Fig.~\ref{fig:3DChannel_severalfields_adapt}b) and at the top of the domain.

>>>>

The computational domain with permeability distribution of 10$^{-4}\le\mathbf{K}\le$10$^{5}$  milidarcy (Fig.~\ref{fig:3DChannel_PermMesh})  was designed to naturally capture preferential flow pathways through channels of large absolute permeability. Pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$ from the left-hand side face. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid, \ie $S_{nw}(t=0)=0.80$, which was continuously displaced by the wetting fluid as the simulation progress. The domain was discretised with 235k tetrahedral \PNDG[1]{1} element-pairs.

As demonstrated in the preliminary model validation (Section~\ref{section:results_initial_model_validation}), crossflow towards the upper region of the initial part of the domain is due to boundary pressures and diffusivities (both at different vertical layers), \ie caused by large permeability gradients (Fig.~\ref{fig:3DChannel_sat}). As the simulation progresses, non-wetting fluid is driven from regions with lower fluid resistance (central channel), although adjacent region (with low permeability) remains stagnant.

Such preferential flow pathways are clearly indicated in Fig.~\ref{fig:3DChannel_planes}, where Darcy velocity vectors overlapped with wetting phase fluid saturation are shown. Fluid displacement occurs mainly in regions of relatively larger permeability, leaving other parts of the domain (\ie regions with low permeability distribution) with little or no momentum, indicating that at these regions the non-wetting fluid (\ie fluid 2) is not effectively displaced.

In Sections \ref{section:results_homo_hete}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces (Fig.~\ref{fig:3DChannel_Isosurf}).

\medskip
A similar numerical simulation (\ie same geometry, initial mesh resolution, initial and boundary conditions) was also performed using dynamically adaptive mesh. During the simulation, mesh resolution ranged from 235k to 643k \PNDG[1]{1} element-pairs (Fig.~\ref{fig:3DChannel_severalfields_adapt}a). Distinct preferable flow pathways and fingers are shown in Fig.~\ref{fig:3DChannel_severalfields_adapt}b-c and~\ref{fig:3DChannel_satvel_XYplanes_adapt}. Saturation of the wetting fluid during the course of the simulation is shown in Fig.~\ref{fig:3DChannel_sat_adapt}. As expected, both simulations (fixed and adaptive mesh, Figs.~\ref{fig:3DChannel_sat} and~\ref{fig:3DChannel_sat_adapt}, respectively) led to similar results, \ie similar flow pathways and fingers development.

However, due to low-order accuracy of element-pairs -- \PNDG[1]{1} instead of \PN[1]{2} of previous sections \citep[for full investigation on numerical accuracy associated with these element-pairs, see][]{salinas2015,salinas_2016,adam_2016, gomes_2017}, and coarser mesh throughout the computational domain, simulation performed with fixed mesh showed larger smeared fluid saturation.
